---
title: "Xenium Learning"
subtitle: "Mouse PFAS"
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: false
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r packages, echo=T, error=FALSE, message=FALSE, warning=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("SeuratDisk")) {install.packages("SeuratDisk"); require("SeuratDisk")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("paletteer")) {install.packages("paletteer"); require("paletteer")} # color palette
if (!require("grDevices")) {install.packages("grDevices"); require("grDevices")} # for grDevices palette
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("tibble")) {install.packages("tibble"); require("tibble")} # for table transformation
#if (!require("geneName")) {install.packages("geneName"); require("geneName")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("ggupset")) {install.packages("ggupset"); require("ggupset")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("viridis")) {install.packages("viridis"); require("viridis")}
if (!require("qs")) {install.packages("qs"); require("qs")}
if (!require("arrow")) {install.packages("arrow"); require("arrow")}

#setRepositories(ind = 1:3, addURLs = c('https://satijalab.r-universe.dev', 'https://bnprks.r-universe.dev/'))
#install.packages(c("BPCells", "presto", "glmGamPoi"))

#library(SeuratData)
library(openxlsx)
library(gplots)
library(ggvenn)
library(BPCells)
library(presto)
library(glmGamPoi)

options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320

```

# Introduction

https://satijalab.org/seurat/articles/seurat5_spatial_vignette_2#mouse-brain-10x-genomics-xenium-in-situ

## Load Xenium Object

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

xenium.obj <- LoadXenium(here("output-XETG00402__0054800__Region_1__20250822__221946"), fov = "fov")

xenium.obj

head(xenium.obj@meta.data)
```

### Visualize Xenium object

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Boundaries(xenium.obj[["fov"]])

ImageDimPlot(xenium.obj, cols = "polychrome", size = 0.75)

```


## Add Clinical information into meta.data

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)

# remove cells with 0 counts
xenium.obj <- subset(xenium.obj, subset = nCount_Xenium > 0)

VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)

VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2)

high_count_cells <- xenium.obj@meta.data %>%
  filter(nCount_Xenium > 2500) %>%
  rownames()

df_export <- data.frame(
  cell_id = high_count_cells,
  group   = "Higher_than_2500_counts"
) %>%
  tibble::rownames_to_column("row") %>%
  select(-row)

# Save as CSV
write.csv(df_export, "cellid_higher_than_2500_counts.csv", row.names = FALSE)

```

Increasing PCA dims to 50

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width = 12, fig.height = 8}

xenium.obj <- SCTransform(xenium.obj, assay = "Xenium")
xenium.obj <- RunPCA(xenium.obj, npcs = 50, features = rownames(xenium.obj))
xenium.obj <- RunUMAP(xenium.obj, dims = 1:50)
xenium.obj <- FindNeighbors(xenium.obj, reduction = "pca", dims = 1:50)
xenium.obj <- FindClusters(xenium.obj, resolution = 2)

DimPlot(xenium.obj, group.by = "SCT_snn_res.2")

```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width = 12, fig.height = 8}

xenium.obj <- qread(here("region1_xenium.qs"))

markers.to.plot1 <- c("Lrp2",         # PTS1, PTS2, PTS3
                      "Slc5a12",      # PTS1
                      "Slc13a3",      # PTS2
                      "Slc16a9",      # PTS3
                      "Epha7",        # DTL
                      "Cryab",        # DTL
                      "Slc12a1",      # TALA, TALB, MD
                      "Cldn16",       # TALA
                      "Cldn10",       # TALB
                      "Nos1",         # MD
                      "Slc12a3",      # DCT1, DCT2
                      "Egf",          # DCT1
                      "Slc8a1",       # CNT, DCT2
                      "Aqp2",         # PC
                      "Slc4a1",       # ICA
                      "Kit",          # ICA
                      "Slc26a4",      # ICB
                      "Nphs1",        # PODO
                      "Ncam1",        # PEC
                      "Upk1b",        # URO
                      "Flt1",         # EC
                      "Emcn",         # EC
                      "Lyve1",        # LYMPH
                      "Kdr",          # LYMPH
                      "Pdgfra",       # FIB
                      "Pdgfrb",       # FIB, MES, VSMC
                      "Piezo2",       # MES
                      "Acta2",        # VSMC
                      "Ptprc",        # LYMPHO, MACRO
                      "Skap1",        # LYMPHO
                      "Cd74"          # MACRO
)

p1 <- DotPlot(xenium.obj,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  labs(x = "Gene Features", y = "Clusters")

p1

```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width = 12, fig.height = 8}

markers.to.plot2 <- c("Lrp2",         # PTS1, PTS2, PTS3
                      "Slc5a12",      # PTS1
                      "Slc13a3",      # PTS2
                      "Slc16a9",      # PTS3
                      "Slc5a2",
                      "Epha7",        # DTL
                      "Cryab",        # DTL
                      "Thbs1",
                      "Slc4a11",
                      "Scin",
                      "Slc12a1",      # TALA, TALB, MD
                      "Cldn16",       # TALA
                      "Cldn10",       # TALB
                      "Nos1",         # MD
                      "Slc12a3",      # DCT1, DCT2
                      "Pvalb",
                      "Egf",          # DCT1
                      "Slc8a1",       # CNT, DCT2
                      "Aqp2",         # PC
                      "Slc4a1",       # ICA
                      "Kit",          # ICA
                      "Slc26a4",      # ICB
                      "Nphs1",        # PODO
                      "Ncam1",        # PEC
                      "Upk1b",        # URO
                      "Flt1",         # EC
                      "Emcn",         # EC
                      "Vwf",
                      "Cdh5",
                      "Tek",
                      "Plvap",
                      "Sox17",
                      "Lyve1",        # LYMPH
                      "Kdr",          # LYMPH
                      "Pdgfra",       # FIB
                      "Col1a2",
                      "Vcan",
                      "Pdgfrb",       # FIB, MES, VSMC
                      "Piezo2",       # MES
                      "Ren1",
                      "Acta2",        # VSMC
                      "Ptprc",        # LYMPHO, MACRO
                      "Igkc",
                      "Dock2",
                      "Skap1",        # LYMPHO
                      "Cd74"          # MACRO
)

p5 <- DotPlot(xenium.obj,
features = markers.to.plot2,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  labs(x = "Gene Features", y = "Clusters")

p5

```





```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

df <- FindAllMarkers(xenium.obj, only.pos = TRUE, min.pct = .25)


# Order by avg_log2FC
all_markers <- df %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, everything())

# Split by cluster (ident column)
marker_list <- split(all_markers, all_markers$cluster)

# Sort names alphanumerically
sorted_cluster_names <- names(marker_list)[order(as.numeric(names(marker_list)))]

# Create a workbook
wb <- createWorkbook()

# Add each cluster as a new worksheet
for (cluster_name in sorted_cluster_names) {
  addWorksheet(wb, sheetName = cluster_name)
  writeData(wb, sheet = cluster_name, marker_list[[cluster_name]])
}

date <- format(Sys.Date(), "%Y%m%d")

# Save workbook
saveWorkbook(wb, here(paste0(date, "_", "FindAllMarkers_By_Cluster.xlsx")), overwrite = TRUE)


```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

library(dplyr)

# Get cell IDs and cluster assignments
df_export <- data.frame(
  cell_id = Cells(xenium.obj),
  group   = Idents(xenium.obj)
) %>%
  tibble::rownames_to_column("row") %>%
  select(-row)

# Save as CSV
write.csv(df_export, "cellid_SCT_snn_res2_cluster_export.csv", row.names = FALSE)

DimPlot(xenium.obj, group.by = "SCT_snn_res.1")

head(xenium.obj@meta.data)

p1 <- DotPlot(xenium.obj,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5,
group.by = "SCT_snn_res.1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  labs(x = "Gene Features", y = "Clusters")

p1

DimPlot(xenium.obj2, group.by = "Cluster") + NoLegend()

p1 <- DotPlot(xenium.obj2,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5,
group.by = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  labs(x = "Gene Features", y = "Clusters")

p1



```

## Translate Xenium Explorer Clusters to Seurat

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

cells <- read_csv("Selection_1_cells_stats.csv", skip = 2) %>%
  as.data.frame() %>%                   # dplyr/tibble prefers explicit cols
  column_to_rownames(var = colnames(.)[1])   # use first column as rownames

xenium.obj <- AddMetaData(xenium.obj, cells)

head(xenium.obj@meta.data)

DimPlot(xenium.obj)

DimPlot(xenium.obj, group.by = "Cluster")

t4 <- table(xenium.obj@meta.data$SCT_snn_res.1, xenium.obj@meta.data$Cluster)

t5 <- t(t4)

prop.t1 <- prop.table(t5, margin = 2) 
prop.t1

t5 <- as.data.frame(prop.t1)
colnames(t5) <- c('Cell_type', 'Group', 'Frequency')

# Original plot

ggplot(t5, aes(fill=Cell_type, y=Frequency, x=Group)) + 
  geom_bar(position="fill", stat = "identity", fun.y = "mean", colour="black") +
  theme_classic()

prop.t2 <- prop.table(t4, margin = 2) 
prop.t2

t4 <- as.data.frame(prop.t2)
colnames(t4) <- c('Cell_type', 'Group', 'Frequency')


ggplot(t4, aes(fill=Cell_type, y=Frequency, x=Group)) + 
  geom_bar(position="fill", stat = "identity", fun.y = "mean", colour="black") +
  theme_classic()

xenium.obj@meta.data <- xenium.obj@meta.data %>%
  mutate(high_count = ifelse(nCount_Xenium > 2500, "yes", "no"))

head(xenium.obj@meta.data)

VlnPlot(xenium.obj, "Area (µm^2)", group.by = "high_count")


xenium.obj@meta.data$high_count <- factor(
  xenium.obj@meta.data$high_count,
  levels = c("yes", "no")
)
DimPlot(xenium.obj, group.by = "high_count", order = T)

unique(xenium.obj@meta.data$segmentation_method)

DimPlot(xenium.obj, group.by = "segmentation_method", order = T)


mosaicplot(table(xenium.obj$SCT_snn_res.1, xenium.obj$segmentation_method))

VlnPlot(xenium.obj, "nCount_Xenium", group.by = "segmentation_method")

```

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width = 12, fig.height = 8}

xenium.obj@meta.data <- xenium.obj@meta.data %>%
  mutate(subclass = dplyr::case_when(
    seurat_clusters == 0  ~ "EC",
    seurat_clusters == 1  ~ "PTS2",
    seurat_clusters == 2  ~ "PTS3",
    seurat_clusters == 3  ~ "PTS1",
    seurat_clusters == 4  ~ "TALA",
    seurat_clusters == 5  ~ "PTS2",
    seurat_clusters == 6  ~ "PTS2",
    seurat_clusters == 7  ~ "PTS2",
    seurat_clusters == 8  ~ "PTS1",
    seurat_clusters == 9  ~ "TALA",
    seurat_clusters == 10 ~ "FIB",
    seurat_clusters == 11 ~ "PTS1",
    seurat_clusters == 12 ~ "MACRO",
    seurat_clusters == 13 ~ "EC",
    seurat_clusters == 14 ~ "TALB",
    seurat_clusters == 15 ~ "FIB",
    seurat_clusters == 16 ~ "PTS2",
    seurat_clusters == 17 ~ "PC",
    seurat_clusters == 18 ~ "PTS3",
    seurat_clusters == 19 ~ "DCT1",
    seurat_clusters == 20 ~ "PC",
    seurat_clusters == 21 ~ "VSMC",
    seurat_clusters == 22 ~ "ICA",
    seurat_clusters == 23 ~ "CNT",
    seurat_clusters == 24 ~ "PTS1",
    seurat_clusters == 25 ~ "FIB",
    seurat_clusters == 26 ~ "TALA",
    seurat_clusters == 27 ~ "DTL",
    seurat_clusters == 28 ~ "MES",
    seurat_clusters == 29 ~ "PTS2",
    seurat_clusters == 30 ~ "PTS1",
    seurat_clusters == 31 ~ "EC",
    seurat_clusters == 32 ~ "DCT1",
    seurat_clusters == 33 ~ "THAL",
    seurat_clusters == 34 ~ "EC",
    seurat_clusters == 35 ~ "THAL",
    seurat_clusters == 36 ~ "ICB",
    seurat_clusters == 37 ~ "PODO",
    seurat_clusters == 38 ~ "PTS2",
    seurat_clusters == 39 ~ "FAT",
    seurat_clusters == 40 ~ "URO",
    seurat_clusters == 41 ~ "DCT2",
    seurat_clusters == 42 ~ "EC",
    seurat_clusters == 43 ~ "EC",
    seurat_clusters == 44 ~ "REN",
    seurat_clusters == 45 ~ "FIB",
    seurat_clusters == 46 ~ "PC",
    seurat_clusters == 47 ~ "MACRO",
    seurat_clusters == 48 ~ "LYMPHO",
    seurat_clusters == 49 ~ "EC",
    seurat_clusters == 50 ~ "PTS2",
    seurat_clusters == 51 ~ "MD",
    seurat_clusters == 52 ~ "VSMC",
    TRUE ~ NA_character_
  ))
DimPlot(xenium.obj, group.by = "subclass", label = T)

# Get cell IDs and cluster assignments
df_export <- data.frame(
  cell_id = Cells(xenium.obj),
  group   = xenium.obj@meta.data$subclass
) %>%
  tibble::rownames_to_column("row") %>%
  select(-row)

# Save as CSV
write.csv(df_export, "cellid_subclass_cluster_export.csv", row.names = FALSE)

df_export <- data.frame(
  cell_id = Cells(xenium.obj),
  group   = xenium.obj@meta.data$seurat_clusters
) %>%
  tibble::rownames_to_column("row") %>%
  select(-row)

# Save as CSV
write.csv(df_export, "cellid_subclass_number_cluster_export.csv", row.names = FALSE)



subclass_order <- c("PTS1", "PTS2", "PTS3", "DTL", "TALA", "TALB", "MD", "DCT1", "DCT2", "CNT", "PC", "ICA", "ICB", "PODO", "PEC", "URO", "EC", "FIB", "MES", "VSMC", "LYMPHO", "MACRO", "X1", "X2", "X3", "X4")

xenium.obj@meta.data$subclass <- factor(xenium.obj@meta.data$subclass, levels = subclass_order)

# Change the identities to align with new subclass names and make umap

Idents(xenium.obj) <- xenium.obj@meta.data$subclass

DimPlot(xenium.obj, reduction = "umap", label = TRUE, label.size = 4) +  
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle(paste0("Subclass Annotation for Xenium"))

p2 <- DotPlot(xenium.obj,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  labs(x = "Gene Features", y = "Clusters")

p2

```


library(stringr)

xenium.obj2@meta.data$Cluster <- factor(
  xenium.obj2@meta.data$Cluster,
  levels = str_sort(unique(xenium.obj2@meta.data$Cluster), numeric = TRUE)
)
```

# Session Info

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Sys.time()

sessionInfo()

```
