---
title: "JWN Effort Sankey"
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("BiocManager")) {install.packages("BiocManager"); require("BiocManager")}
if (!require("tibble")) {install.packages("tibble"); require("tibble")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} #color
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("plotly")) {install.packages("plotly"); require("plotly")}
if (!require("gginnards")) {install.packages("gginnards"); require("gginnards")}


#remotes::install_github("davidsjoberg/ggsankey")
library(ggsankey)
library(tidyr)

here()

```


#Introduction

The goal of this project is to create a Sankey diagram that shows where my effort and output go as a research scientist. There will goal is to show revenue input followed by output in the form of research, teaching, and service. For the first time with this plot, I'm going to try to code it so that I can calculate the ratio of effort that goes into teach category. 

# Research Grant Inputs

## Load Data

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

df <- read.xlsx(here("Dataset", "Revenue Input.xlsx"))

df

```
## Grants

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}


df_subcontract <- df %>%
  uncount(Subcontract.value, .remove = FALSE) %>%
  mutate(Subcontract_Number = Subcontracts) %>% 
  select(Subcontracts)


df_Grants <- df %>%
  uncount(Grants.value, .remove = FALSE) %>%
  mutate(Grants.Number = Grants) %>% 
  select(Grants)

df1 <- df_subcontract
df2 <- df_Grants

max_rows <- max(nrow(df1), nrow(df2))

df1_extended <- df1 %>% 
  bind_rows(tibble(Subcontracts = rep(NA, max_rows - nrow(df1))))

df2_extended <- df2 %>% 
  bind_rows(tibble(Grants = rep(NA_character_, max_rows - nrow(df2))))

df_merged <- cbind(df1_extended , df2_extended)

df_merged$Corporeal <- "JWN"

```

# Make SanKey

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=10}

df3 <- df_merged %>% make_long(Subcontracts, Grants, Corporeal) %>% 
  filter(!is.na(node))

df3.input <- df3

p1 <- ggplot(df3, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE)

p1 <- p1 + theme_classic() + theme(legend.position = "none")



p1

p_plotly <- ggplotly(p1, tooltip = "text")

p_plotly

```



## Research Outputs

### Load Data

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

df <- read.xlsx(here("Dataset", "Research Output.xlsx"))

df

```

# Start with Main Level

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

df[is.na(df)] <- 0

df_Main <- df %>%
  uncount(Main.value, .remove = FALSE) %>%
  mutate(Main_Number = Main) %>% 
  select(Main)

df3 <- df_Main  
df3$Corporeal <- "JWN"

# Make Corporeal the first column using dplyr

df3 <- df3 %>% select(Corporeal, Main)

```


```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}


#df3 <- df3 %>%  tidyr::expand(Corporeal, Main) %>% 
#  filter(!is.na(Main))

df3 <- df3 %>% make_long(Corporeal, Main) %>% 
  filter(!is.na(node))

df4 <- rbind(df3.input, df3)

df4 <- df4 %>% filter(!(x == "Corporeal" & is.na(next_x)))

p1 <- ggplot(df4, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE)

p1


```

## Add Research Level 1

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

research_value <- df %>%
  filter(Main == "Research") %>%
  pull(Main.value) %>% 
  `/`(100)

df_Research.Level1 <- df %>%
    uncount(Research.Level1.value * research_value, .remove = FALSE) %>%
  mutate(Level1 = Research.Level1) %>% 
  select(Level1)

df3 <- df_Research.Level1  
df3$Main <- "Research"

df3 <- df3 %>% select(Main, Level1)

df3 <- df3 %>% make_long(Main, Level1) %>% 
  filter(!is.na(node))

df4 <- rbind(df4, df3)

df5 <- df4 %>% filter(!(x == "Main" & node == "Research" & is.na(next_x)))

p1 <- ggplot(df5, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE)

p1


```


## Add Service Level 1

Needs to be dividable by the service value (e.g. 5 didn't work, but 10 did becuase it was divisable to 1% total effort) 

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

service_value <- df %>%
  filter(Main == "Service") %>%
  pull(Main.value) %>% 
  `/`(100)

df[is.na(df)] <- 0

df_Service.Level1 <- df %>%
  uncount(Service.Level1.value * service_value, .remove = FALSE) %>%
  mutate(Level1 = Service.Level1) %>% 
  select(Level1)

df3 <- df_Service.Level1  
df3$Main <- "Service"

df3 <- df3 %>% select(Main, Level1)

df3 <- df3 %>% make_long(Main, Level1) %>% 
  filter(!is.na(node))

df5 <- rbind(df5, df3)

df6 <- df5 %>% filter(!(x == "Main" & node == "Service" & is.na(next_x)))

p1 <- ggplot(df6, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE
                      ,type = "sankey")

p1

```
## Add Teaching Level 1

Needs to be dividable by the service value (e.g. 5 didn't work, but 10 did becuase it was divisable to 1% total effort) 

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

df

Teaching_value <- df %>%
  filter(Main == "Teaching and Mentorship") %>%
  pull(Main.value) %>% 
  `/`(100)

df[is.na(df)] <- 0

df_Teaching.Level1 <- df %>%
  uncount(Teaching.Level1.value * Teaching_value, .remove = FALSE) %>%
  mutate(Level1 = Teaching.Level1) %>% 
  select(Level1)

df3 <- df_Teaching.Level1  
df3$Main <- "Teaching and Mentorship"

df3 <- df3 %>% select(Main, Level1)

df3 <- df3 %>% make_long(Main, Level1) %>% 
  filter(!is.na(node))

df6 <- rbind(df6, df3)

df7 <- df6 %>% filter(!(x == "Main" & node == "Teaching and Mentorship" & is.na(next_x)))

p1 <- ggplot(df6, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE
                      ,type = "sankey")

p1

```

## Add Collab Level 2

Needs to be an integer of research_value x collab_value = .8 * .5 = .4

Multiply column in excel by .4 and make sure that it's a whole number.

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

Collab_value <- df %>%
  filter(Research.Level1 == "Collaborative") %>%
  pull(Research.Level1.value) %>% 
  `/`(100)

df[is.na(df)] <- 0

df_Collab.Level2 <- df %>%
  uncount(Collab.Level2.value * Collab_value * research_value, .remove = FALSE) %>%
  mutate(Level2 = Collab.Level2) %>% 
  select(Level2)

df3 <- df_Collab.Level2  
df3$Level1 <- "Collaborative"

df3 <- df3 %>% select(Level1, Level2)

df3 <- df3 %>% make_long(Level1, Level2) %>% 
  filter(!is.na(node))

df8 <- rbind(df7, df3)

df9 <- df8 %>% filter(!(x == "Level1" & node == "Collaborative" & is.na(next_x)))

p1 <- ggplot(df9, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE
                      ,type = "sankey")

p1

p_plotly <- ggplotly(p1, tooltip = "text")

p_plotly


```

## Add Primary Level 2

Needs to be an integer of research_value x collab_value = .8 * .5 = .4

Multiply column in excel by .4 and make sure that it's a whole number.

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=10, fig.height=10}

df9 <- df8 %>% filter(!(x == "Level1" & node == "Collaborative" & is.na(next_x)))

Primary_value <- df %>%
  filter(Research.Level1 == "Primary") %>%
  pull(Research.Level1.value) %>% 
  `/`(100)

df[is.na(df)] <- 0

df_Research.Level2 <- df %>%
  uncount(Research.Level2.value * Primary_value * research_value, .remove = FALSE) %>%
  mutate(Level2 = Research.Level2) %>% 
  select(Level2)

df3 <- df_Research.Level2  
df3$Level1 <- "Primary"

df3 <- df3 %>% select(Level1, Level2)

df3 <- df3 %>% make_long(Level1, Level2) %>% 
  filter(!is.na(node))

df10 <- rbind(df9, df3)

df11 <- df10 %>% filter(!(x == "Level1" & node == "Primary" & is.na(next_x)))

p1 <- ggplot(df11, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE
                      ,type = "sankey")

p1

p_plotly <- ggplotly(p1, tooltip = "text")

p_plotly

```



## Factor nodes in correct order

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}
df.test <- df11 

df.test$node <- as.factor(df.test$node)

levels(df.test$node)

levels_to_assign <- c("Emery R01",
                      "Gurley R01",
                      "Ellison R01",
                      
                      "K01",
                      "Startup",
                      "Subcontracts",
                     
                       "JWN",
                      
                      "Research",
                      "Service",
                      "Teaching and Mentorship",
                      
                      "Collaborative",
                      "Primary",
                      
                      "Committees",
                      "Peer Review",
                      "Journal Club",
                      "Teaching",
                      "Mentorship",
                      "Course Development",
                      
                      "Nakai",
                      "Chung",
                      "Srivastava",
                      "Duvoisin",
                      "McDonough",
                      "Gurley",
                      "Hutchens",
                      "Ellison",
                      "Emery",
                      
                      
                      "Kidney-Specific Drugs",
                      "PFAS and Kidney Disease",
                      "Pdgfrb-DREADD",
                      "Pdgfrb-INTACT",
                      "Pdgfrb-AT1aR KO"
                      
                      )   

                      
df.test$node <- factor(df.test$node, levels = levels_to_assign)


p1 <- ggplot(df.test, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = factor(node)
                     , label = node)
             )

p1 <- p1 +geom_sankey(flow.alpha = 0.5
                      , node.color = "black"
                      ,show.legend = FALSE)


p1 <- p1 + theme_classic() + theme(legend.position = "none")

p1

p_plotly <- ggplotly(p1, tooltip = "text")

p_plotly

```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=14, fig.height=6}

p1 <- p1 + theme_classic() + theme(legend.position = "none") +
  geom_sankey_label(size = 3.5, color = 1, fill = "white")

# remove axis lines and axis labels

p1 <- p1 + theme(axis.line = element_blank(),
                 axis.text.x = element_blank(),
                 axis.text.y = element_blank(),
                 axis.ticks = element_blank(),
                 axis.title.x = element_blank(),
                 axis.title.y = element_blank())


p2 <- p1 + scale_fill_viridis_d(direction = -1, option = "viridis") + 
      scale_colour_viridis_d(direction = -1, option = "viridis")
p2

p3 <- p1 + scale_fill_viridis_d(direction = 1, option = "viridis") + 
      scale_colour_viridis_d(direction = 1, option = "viridis")
p3

p4 <- p1 + scale_fill_viridis_d(direction = -1, option = "turbo") + 
      scale_colour_viridis_d(direction = -1, option = "turbo")
p4


```

# Final Sankey Plot

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=14, fig.height=6}


p2 <- p1 + scale_fill_viridis_d(direction = -1, option = "viridis") + 
      scale_colour_viridis_d(direction = -1, option = "viridis") +  
      annotate("text", x = 2, y = 200, label = "Research Revenue", color = "black", size = 7) +
      annotate("text", x = 4, y = 200, label = "Research Output", color = "black", size = 7) 

p2

```

## Exploring Figure Data

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=14, fig.height=6}

data <- ggplot_build(p1)
summary(data$plot)


```

```{r load data, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=14}

sessionInfo()

```


  